---
title: "Public Transport Usage in Victoria"
subtitle: "Which modes are most used, and where should investment be focused?"
author:
  - Vishal Sampath Kumar
  - Rikhitha Varma
  - Socheaneath Hour
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    incremental: false
    smaller: true
    title-slide-attributes:
      data-background-color: "#f7f9fb"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false

library(tidyverse)
library(janitor)
library(readr)
library(knitr)
library(ggplot2)
library(scales)
library(stringr)
library(lubridate)

monthly_clean <- read_csv("data/processed/monthly_clean.csv")
stations_clean <- read_csv("data/processed/stations_clean.csv")

top_stations <- stations_clean |>
  arrange(desc(pax_annual))

mode_summary <- monthly_clean |>
  group_by(mode) |>
  summarise(total_boardings = sum(boardings, na.rm = TRUE), .groups = "drop")
```

## Problem and questions

:::: {.columns}
::: {.column width="55%"}
### Why this matters

- Public transport supports mobility, access, and economic activity across Victoria.
- Investment decisions should be guided by observed passenger demand.
- Understanding demand by **mode** and **location** helps target upgrades where they will have the greatest effect.
:::

::: {.column width="45%"}
### Research questions

1. Which public transport modes are most used in Victoria?
2. Which metropolitan train stations have the highest demand?
3. Where should future investment be prioritised?
:::
::::

## Datasets

:::: {.columns}
::: {.column width="50%"}
### Dataset 1: Monthly patronage

- Victorian public transport boardings
- Monthly observations from 2018 to 2025
- Includes:
  - Metropolitan train
  - Metropolitan tram
  - Metropolitan bus
  - Regional train
  - Regional bus
  - Regional coach
:::

::: {.column width="50%"}
### Dataset 2: Station entries

- FY2024-25 metropolitan train station entries
- Annual passenger entries by station
- Used to identify highest-demand stations in Melbourne's network
:::
::::

## Data preparation and methods

:::: {.columns}
::: {.column width="50%"}
### Data preparation

- Standardised column names
- Converted passenger counts to numeric values
- Removed formatting inconsistencies
- Reshaped monthly patronage data from wide to long format
- Checked for missing values
:::

::: {.column width="50%"}
### Analysis approach

- Summed total boardings by mode across the study period
- Ranked stations by annual passenger entries
- Visualised:
  - Total patronage by mode
  - Monthly patronage trends over time
:::
::::

## Results: Mode usage

:::: {.columns}
::: {.column width="38%"}
### Key finding

- **Metropolitan train** is the most used public transport mode in Victoria.
- It is followed by metropolitan tram and metropolitan bus.
- Regional services carry much smaller totals across the study period.

### Interpretation

Rail investment is likely to affect the largest number of passengers overall.
:::

::: {.column width="62%"}
```{r}
#| fig-width: 8
#| fig-height: 4.5
#| out-width: "100%"

mode_summary |>
  mutate(mode = str_replace_all(mode, "_", " ") |> str_to_title()) |>
  ggplot(aes(
    x = reorder(mode, total_boardings),
    y = total_boardings
  )) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    x = NULL,
    y = "Total boardings"
  ) +
  theme_minimal(base_size = 16)
```
:::
::::

## Results: Busiest stations

:::: {.columns}
::: {.column width="42%"}
### Highest-demand stations

- Flinders Street
- Southern Cross
- Melbourne Central
- Parliament
- Richmond

These stations stand out as major demand hubs and are strong candidates for capacity and accessibility upgrades.
:::

::: {.column width="58%"}
```{r}
top_stations |>
  head(5) |>
  select(Station = stop_name, `Annual entries` = pax_annual) |>
  mutate(`Annual entries` = comma(`Annual entries`)) |>
  knitr::kable()
```
:::
::::

## Results: Patronage trends

:::: {.columns}
::: {.column width="35%"}
### Trend over time

- Patronage dropped sharply in 2020 during COVID-19.
- Recovery occurred across all modes after the pandemic period.
- Metropolitan trains and trams remained the dominant modes throughout.

### Insight

Travel demand recovered, but disruption changed usage patterns over time.
:::

::: {.column width="65%"}
```{r}
monthly_clean |>
  mutate(
    date = make_date(year, month, 1),
    mode = str_replace_all(mode, "_", " ") |> str_to_title()
  ) |>
  ggplot(aes(
    x = date,
    y = boardings,
    colour = mode
  )) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = comma) +
  labs(
    x = NULL,
    y = "Monthly boardings",
    colour = "Mode"
  ) +
  theme_minimal(base_size = 13)
```
:::
::::

## Key findings

| Finding | Evidence | Implication |
|---|---|---|
| Metropolitan train is the most used mode | Highest total boardings across the study period | Rail improvements can benefit the most passengers |
| Trams and buses also carry high demand | Metropolitan tram and bus rank next | Investment should remain multimodal |
| Demand is concentrated at major stations | Top entries occur at central and interchange stations | Station upgrades should focus on high-volume hubs |
| COVID-19 disrupted usage patterns | Clear decline in 2020 followed by recovery | Planning should account for changing travel behaviour |

## Recommendations

- Prioritise **metropolitan rail investment**, especially where high passenger volumes create crowding and reliability risks.
- Target the busiest stations for upgrades such as capacity expansion, accessibility improvements, wayfinding, and passenger flow management.
- Continue support for tram and bus networks, especially where they complement rail and improve first-mile/last-mile access.
- Use ongoing patronage monitoring to update investment priorities as travel patterns continue to recover and evolve.

## Conclusion

- Public transport demand in Victoria is led by metropolitan rail, followed by trams and buses.
- The busiest metropolitan stations should be focal points for near-term infrastructure investment.
- A balanced strategy is strongest: invest heavily in rail hubs while maintaining strong multimodal network support.